|

### 2024-03-15 manuscript/manuscript.qmd
manuscript updates

[The Adolescent Brain Cognitive Development (ABCD) Study®](https://abcdstudy.org/ "Visit the ABCD Study website"){#abcdwebsite} is the largest long-term investigation of neurodevelopment and child health in the United States. Conceived and initiated by the National Institutes of Health (NIH), this landmark prospective longitudinal study aims to transform our understanding of the genetic and environmental factors impacting neurodevelopment and their roles in behavioral and health outcomes across ten years of adolescence (Volkow et al. 2018). At its heart, the study is designed to chart the course of human development across multiple interacting domains from late childhood to early adulthood and to identify factors that lead to both positive and negative outcomes. Central to achieving these goals is the commitment of the ABCD Study® and its NIH funders to an open science framework, intended to facilitate sharing of data and analytical methods, by espousing practices that increase access, integrity, and reproducibility of scientific research. In this sense, the ABCD Study is a collaboration with the larger research community.

The size and scope of the ABCD Study data allow the research community to perform a large variety of developmental analyses of both substantive and methodological interest,  presenting a unique opportunity to significantly advance our understanding of how a multitude of biopsychosocial processes unfold across critical periods of development. In this paper, we describe models and methods for longitudinal analysis of ABCD Study data that can address these fundamental scientific aims, including: 1) characterization of the genetic and environmental factors associated with variation in developmental trajectories; 2) assessment of how the level and timing of exposures may impact subsequent neurodevelopment; 3) quantification of how variation in developmental domains may be associated with outcomes, including mediation models and reciprocal relationships. We instantiate these longitudinal analyses in worked examples using the ABCD Release 5.0 data with accompanying R scripts. Worked examples are available in Quarto files, accessible at **XXXX**.

@@

The ABCD Study enrolled a cohort of n=11,880 participants born between 2006-2008 and aged between 9-11 years at baseline, as well as their parents/guardians. The study sample was recruited from household populations in defined catchment areas for each of the 21 (originally 22) study sites across the United States. Information regarding funding agencies, recruitment sites, investigators, and project organization can be obtained at https://abcdstudy.org/. The ABCD Study design is described in more detail in @garavan2018 and @dick2021.

The ABCD Study is currently collecting longitudinal data on a rich variety of outcomes that will enable the construction of complex statistical models potentially incorporating factors from many domains. Each new wave of data collection provides another building block for characterizing developmental trajectories and implementing longitudinal analyses that allow researchers to characterize normative development, to identify variables that presage deviations from normative development, and to assess a range of variables associated with biopsychosocial outcomes of interest. These data include: 1) a neurocognitive battery [@luciana2018a; @thompson2019]; 2) mental and physical health assessments [@barch2018]; 3) measures of culture and environment [@gonzalez2021; @zucker2018]; 4) substance use [@lisdahl2021]; 5) gender identity and sexual health [@potter2022]; 6) biospecimens [@uban2018]; 7) structural and functional brain imaging [@casey2018; @hagler2019]; 8) geolocation-based environmental exposure data [@fan2021]; 9) wearables, and mobile technology [@bagot2018]; and 10) whole genome genotyping [@loughnan2020]. Many of these measures are collected at in-person annual visits, with brain imaging collected at baseline and every other year going forward. A limited number of assessments are collected in semi-annual telephone interviews between in-person visits.

Data are publicly released roughly annually, currently through the NIMH Data Archive (NDA). By necessity, the study’s earliest data releases consisted primarily of one or two visits per participant. However, the most recent public release as of the writing of this paper (Release 5.0) contains data collected across five annual visits, including three brain imaging assessments (baseline, year 2 follow-up, and year 4 follow-up visits) for at least a subset of the cohort. Hence, starting with Release 5.0, it is feasible for researchers to begin focusing on the estimation of neurodevelopmental and other trajectories.

@@

-   Identify fundamental concepts

@@

-   Highlight key challenges

@@

There are several important concepts to consider when conducting longitudinal analyses in a developmental context. These include different ways of thinking about developmental course, whether certain periods of development are relatively sensitive or insensitive to various types of insults or stressors, whether some time periods or situations inhibit the expression of individual differences due to extreme environmental pressures, and whether the same behavior manifested at different times represents the same or different phenomena.

@@

For example, collecting two or more within-person observations on the same construct at different times enables estimation of individual rates of change (slopes); more observations allow for more precise estimates of individual slopes, as well as characterization of non-linear development. Rate of change or other trajectory characteristics may be more informative about individuals than the simple snapshots of level differences that cross-sectional data are limited to informing. Cross-sectional age-related differences across individuals are poor substitutes for longitudinal trajectory estimates except under highly restrictive assumptions, e.g., parallel trajectories and lack of age cohort and experience effects [@thompson2011]. Appreciation of these and other issues can help to guide the analysis and interpretation of data and aid translation to clinical and public health applications.

@@

Adolescent development progresses normatively from less mature to more mature levels of functioning. However, unique epochs and experiences can alter the course of this idealized form of development. Consider research that shows cannabis use during adolescence is associated with later psychosis to a greater degree than cannabis use initiated later in development [@arseneault2002; @bechtold2016; @hasan2020; @semple2005]; or similarly, experimental research on rodents that shows rodent brains to be especially sensitive to the neurotoxic effects of alcohol on brain structure and learning early in development, corresponding to early adolescence in humans [@spear2016; @crews2000; @ji2018]. In another example, longitudinal data from the National Consortium on Alcohol in Adolescence (NCANDA) show that binge drinking is associated more strongly with decrements in gray matter volume early in adolescence compared to later [@infante2022]. These examples highlight the importance of considering the role of vulnerable periods – e.g., temporal windows of rapid brain development or remodeling during which the effects of environmental stimuli on the developing brain may be particularly pronounced– when trying to establish an accurate understanding of the association between exposures and outcomes.

@@

Whereas vulnerable periods heighten neurobiological susceptibility to environmental influences, at other times, environmental exposures will tend to suppress stability and disrupt the orderly stochastic process of normative development [e.g., @schulenberg2019]. This situation reflects a developmental disturbance in that the normal course of development is “altered” for a time by some time-limited process. In such cases, we might find that prediction of behavior in the period of the disturbance is reduced and/or, similarly, the behavior exhibited during the disturbance might have less predictive power with respect to distal outcomes compared to the behavior exhibited before and following the disrupted period. That is, once the environmental presses are removed (or the individual is removed from the environment), patterns of individual differences (and autoregressive effects) recover to levels similar to those prior to entering the environment. For example, in @infante2022, recent binge drinking appears to be most predictive of gray matter volume trajectories, as opposed to more distal prior binge drinking or cumulative number of binge drinks, suggesting the potential for recovery of gray matter trajectories to prior levels of growth if binge drinking levels subside.

@@

Normative development can also be upended by experiences (e.g., drug use) that, through various mechanisms, disrupt the normal flow of development wherein each stage establishes a platform for the next. For instance, substance use could lead to association with deviant peers, precluding opportunities for learning various adaptive skills and prosocial behaviors, in effect creating a “snare” that delays psychosocial development, such as maturing out of adolescent antisocial behavior [@moffitt2015]. Relatedly, the consequences of these types of events can cascade (e.g., school dropout, involvement in the criminal justice system) so that the effects of the snare are amplified [e.g., @masten2005; @rogosch2010]. Although conceptually distinct from vulnerable periods, both types of developmental considerations highlight the importance of viewing behavior in the context of development and attempting to determine how various developmental pathways unfold. Longitudinal data are crucial in this context to assess individual levels of development prior to and following onset of experiences or other environmental factors; e.g., the ABCD Study collected data starting at ages 9-10 and hence before the onset of substance use for the vast majority of participants.

**2.1.4 Intermediate processes and feedback loops**
**Something here on the concept of mediation and cross-lagged relationships?**

@@

## Interpretation / Issues / Pitfalls & Assumption

The hallmark characteristic of longitudinal data analysis is the administration of repeated measurements of the same constructs on assessment targets (e.g., individuals, families) across time. The primary rationale for collecting longitudinal data is to assess within-person change over time, allowing researchers to estimate individual developmental trajectories and the genetic and/or environmental factors that may impact these trajectories. Administering repeated measurements more frequently or over longer time spans enables researchers to ask more nuanced questions and to make stronger inferences.

@@

Although the clear leap from cross-sectional to the realm of longitudinal data involves going from one assessment to two or more assessments, there are also notable distinctions in designs based on two-assessment points versus three or more measurement occasions. Just as cross-sectional data can be informative in some situations, two waves of data can be beneficial in contexts such as when experimental manipulation is involved (e.g., pre/post tests), or if the central goal is prediction (e.g., trying to predict scores on Variable A at time T as a function of prior scores on Variable A and Variable B at time T-1). At the same time, data based on two assessments are inherently limited on multiple fronts. As @rogosa1982 noted over forty years ago, “Two waves of data are better than one, but maybe not much better” (p. 744).

These sentiments are reflected in more contemporary recommendations regarding best-practice guidelines for prospective data, which increasingly emphasize the benefits of additional measurement occasions for trajectory estimation, model identification and accurate parameter inferences. It is also consistent with research recommending that developmental studies include three or more assessment points, given it is impossible for data based on two-time points to determine the shape of development (given that linear change is the only estimable form for two assessments waves; [see @duncan2009]). Research designs that include three (but preferably more) time points allow for non-linear trajectory estimation and increasingly nuanced analyses that more adequately tease apart sources of variation and covariation among the repeated assessments [@king2018]– a key aspect of developmental research.

To illustrate, developmental theories are useful for understanding patterns of within-individual change over time (discussed in further detail, below); however, two data points provide meager information on change at the person level. This point is further underscored in a recent review of statistical models commonly touted as distinguishing within-individual vs between-individual sources of variance in which the study authors concluded “… researchers are limited when attempting to differentiate these sources of variation in psychological phenomenon when using two waves of data” and perhaps more concerning, “…the models discussed here do not offer a feasible way to overcome these inherent limitations” [@littlefield2021]. It is important to note, however, that despite the current focus on two-wave designs versus three or more assessment waves, garnering three assessment points is not a panacea for longitudinal modeling. Indeed, several contemporary longitudinal models designed to isolate within-individual variability (e.g., the Latent Curve Model with Structured Residuals; @curran2014) require at least four assessments to parameterize fully and, more generally, increasingly accurate and nuanced parameter estimates are obtained as more assessment occasions are used [@duncan2009].

@@

If one were to try to sum up what developmental trajectories in a living organism are exactly, one could plausibly argue they are the patterns of stability and change in its phenotypes as the organism traverses the life course. Symbolically, developmental trajectories can be expressed as fi(t), a possibly multivariate function of time t, specific to the ith individual and typically taking values in the real numbers for continuous phenotypes and the integers for discrete phenotypes. Ideally, t is a biologically meaningful temporal index (e.g., calendar age) as opposed to an exogenous progression of events (e.g., study visit number). Properties of interest might include rate of change over time, degree of smoothness (e.g., continuously differentiable), shape (e.g., polynomial or asymptotic behavior), how and how much fi(t) differs across individuals, and what factors predict either within-individual variation (at different times) or between-individual variation (either overall or at specific times).

There are a few different ways to think about patterns of stability and change (see Figure 1). Consider measuring school disengagement at the start of middle school and the end of middle school . A common first step may be to compare sixth graders’ average disengagement values and eighth graders’ disengagement values. This comparison of the average scores for the same group of individuals at multiple time points is referred to as “mean-level”, as it provides information about change over time (or lack thereof) for an outcome of interest aggregated across members of a group. In contrast, “between-individual” stability could be assessed, e.g., by calculating the Spearman correlation between the values obtained at different time points (e.g., ‘disengagement in sixth grade’ with ‘disengagement in eighth grade). This analysis focuses on the degree to which individuals retain their relative placement in a group across time. Consider someone who reported the lowest frequencies of disengagement in 6th grade and may report significantly higher disengagement over middle school (i.e., exhibit high levels of change), but report the lowest frequencies of disengagement in eighth grade. That is, the individual is manifesting rank-order stability, even in the context of high mean-level change.

Both types of stability and change are important. Mean-level change in certain traits might help to explain why, in general, populations of individuals tend to be particularly vulnerable to the effects of environmental factors in specific age ranges; rank-order stability might help to quantify the extent to which certain characteristics of the individual are more or less trait-like compared to others. For example, in some areas of development, considerable mean-level change occurs over time (e.g., changes in Big 5 personality traits; @bleidorn2022), but exhibit relatively high rank-order stability, at least over shorter measurement intervals [@bleidorn2022; @roberts2000; @roberts2006]. Despite the useful information afforded by examining mean-level and rank-order stability and change, these approaches are limited in that they provide little information about the overall patterns of within-individual change (i.e., trajectories) and, in turn, can result in fundamental misinterpretations about substantial or meaningful changes in an outcome of interest [@curran2011]. For example, questions related to the impact of early-onset substance use on brain development focus on changes within a given individual (i.e., intraindividual differences). The longitudinal nature of the ABCD Study, which will provide researchers with over ten time points for some constructs (e.g., substance use) across a ten-year period, allows for the study of within-person processes.

@@

### Use of appropriate longitudinal models
There is growing recognition that statistical models commonly applied to longitudinal data often fail to align with the developmental theory they are being used to assess [e.g., @curran2011; @hoffman2015; @littlefield2021]. First, developmental studies typically involve the use of prospective data to inform theories that are concerned with clear within-person processes (e.g., how phenotypes change or remain stable within individuals over time, [e.g., @curran2011]). Despite this, methods generally unsuited for disaggregating between- and within-person effects (e.g., cross-lagged panel models [CLPM]) remain common within various extant literatures. Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time [see @littlefield2021; @orth2021]. Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual’s mean (or trait) level across time (e.g., random-intercept cross-lagged panel model [RI-CLPM], @hamaker2015; latent curve models with structured residuals [LCM-SR], @curran2014). It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context (see @littlefield2022, for further discussion).

Second, many statistical models assume certain characteristics about the data to which they are being applied. Common assumptions of parametric statistical models (e.g., linear mixed-effects models) include normality and equality of variances. These assumptions must be carefully considered before finalizing analytical approaches, so that valid inferences can be made from the data, as violation of a model’s assumptions can substantively invalidate the interpretation of results. For example, longitudinal data can exhibit heterogeneous variability (i.e., the variance of the response changes over the duration of the study) that may need to be accounted for within a model. Another pertinent modeling assumption is whether trajectories are linear or non-linear. With 2-3 assessments per individual, only a linear model of within-person change is usually feasible; with more time points, higher level polynomials or models with more flexible trajectory shapes (e.g., curve smoothing via cubic splines) becomes possible [REFs]. Note, however, baseline age in the ABCD Study ranges over two full years; for some outcomes it may be desirable to include a (possibly nonlinear) effect of baseline age along with a linear effect of within-person change in age with only 2-3 assessment times [@thompson2013]. As the study progresses and more times points are assessed, nonparametric curve estimation at the mean [GAMMs] and at the individual level may also become useful [@ramsay2002].

### Continuous and Discrete Outcomes
Implementing valid and efficient statistical models requires an understanding of the type of data being analyzed. For example, repeated assessments within the ABCD Study can be based on continuous or discrete measures. Examples of discrete measures include repeated assessments of binary variables (e.g., past 12-month alcohol use disorder status measured across ten years), ordinal variables (e.g., caregiver-reported items measuring emotional and behavioral concerns via the Child Behavior Checklist including the categories of “Not True”, “Somewhat True”, and “Very True”), and count variables (e.g., number of cigarettes smoked per day). In many ways, the distributional assumptions of indicators used in longitudinal designs mirror the decision points and considerations when delineating across different types of discrete outcome variables, a topic that spans entire textbooks [e.g., see @lenz2016]. For example, the Mplus manual [@muthen2017] includes examples of a) censored and censored-inflated models, b) linear growth models for binary or ordinal variables, c) linear growth models for a count outcome assuming a Poisson model, and d) linear growth models for a count outcome assuming a zero-inflated Poisson model. Beyond these highlighted examples, other distributions (e.g., negative binomial) can be assumed for the indicators when modeling longitudinal data [@ren2022]. These models can account for issues that can occur when working with discrete outcomes, including overdispersion [when the variance is higher than would be expected based on a given distribution; see @lenz2016]. Given the sheer breadth of issues relevant to determining better models for discrete outcomes, it is not uncommon for texts on longitudinal data analysis to only cover models and approaches that assume continuous variables [e.g., @little2013]. However, some textbooks on categorical data analysis provide more detailed coverage of the myriad issues and modeling choices to consider when working with discrete outcomes [e.g., @lenz2016, Chapter 11 for matched pair/two-assessment designs; Chapter 12 for marginal and transitional models for repeated designs, such as generalized estimating equations, and Chapter 13 for random effects models for discrete outcomes].

>>>>

[The Adolescent Brain Cognitive Development (ABCD) Study®](https://abcdstudy.org/ "Visit the ABCD Study website"){#abcdwebsite} is the largest longitudinal investigation of neurodevelopment and child health in the United States. Conceived and initiated by the National Institutes of Health (NIH), this landmark prospective longitudinal study aims to transform our understanding of the genetic and environmental factors impacting neurodevelopment and their roles in behavioral and health outcomes across ten years of adolescence (@volkow2018). At its heart, the study is designed to chart the course of human development across multiple interacting domains from late childhood to early adulthood and to identify factors that lead to both positive and negative outcomes. Central to achieving these goals is the commitment of the ABCD Study© and its NIH funders to an open science framework, intended to facilitate sharing of data and analytical methods by espousing practices that increase access, integrity, and reproducibility of scientific research. In this context, the ABCD Study is a collaboration with the broader research community.

The size and scope of the ABCD Study data allow the research community to perform a large variety of developmental analyses of both substantive and methodological interest, presenting a unique opportunity to significantly advance our understanding of how a multitude of biopsychosocial processes unfold across critical periods of development. In this paper, we describe models and methods for longitudinal analysis of ABCD Study data that can address these fundamental scientific aims, including: 1) characterization of the genetic and environmental factors associated with variation in developmental trajectories; 2) assessment of how the level and timing of exposures may impact subsequent neurodevelopment; 3) quantification of how variation in developmental domains may be associated with outcomes, including mediation models and reciprocal relationships. We instantiate these longitudinal analyses in worked examples using the ABCD Release 5.0 data with accompanying R scripts. Worked examples are available in Quarto files, accessible in the project's [GitHub repository]( https://github.com/OpenDevSci/LongDev-ABCD/tree/main/examples)

@@

The ABCD Study enrolled a cohort of n=11,880 participants born between 2006-2008 and aged between 9-10 years at baseline, each with a parent/guardian. The study sample was recruited from household populations in defined catchment areas for each of the 21 (originally 22) study sites across the United States. Information regarding funding agencies, recruitment sites, investigators, and project organizations can be obtained at https://abcdstudy.org. The ABCD Study design is described in more detail in @garavan2018 and @dick2021.

The ABCD Study is currently collecting longitudinal data on a rich variety of outcomes that will enable the construction of complex statistical models, potentially incorporating factors from many domains. Each new wave of data collection provides another building block for characterizing developmental trajectories and implementing longitudinal analyses that allow researchers to characterize normative development, to identify variables that presage deviations from normative development, and to assess a range of variables associated with biopsychosocial outcomes of interest. These data include: 1) a neurocognitive battery (@luciana2018; @thompson2019); 2) mental and physical health assessments (@barch2018); 3) measures of culture and environment (@gonzalez2021; @zucker2018); 4) substance use (@lisdahl2021); 5) gender identity and sexual health (@potter2022); 6) biospecimens (@uban2018); 7) structural and functional brain imaging (@casey2018; @hagler2019; @palmer2022); 8) geolocation-based environmental exposure data (@fan2021); 9) wearables and mobile technology (@bagot2018); and 10) whole-genome genotyping (@loughnan2020). Many of these measures are collected at in-person annual visits, with brain imaging collected at baseline and every other year going forward. A limited number of assessments are collected in semi-annual brief telephone or online assessments.

Data are publicly released approximately annually, currently through the NIMH Data Archive (NDA). The study’s earliest data releases consisted primarily of one or two visits per participant. However, the most recent public release as of the writing of this paper (Release 5.1) contains data collected across five annual visits, including three brain imaging assessments (baseline, year 2 follow-up, and year 4 follow-up visits) for at least a subset of the cohort. Hence, starting with Release 5.0, it is feasible for researchers to begin focusing on the characterization of neurodevelopmental and other trajectories.

@@

-   Fundamental concepts

@@

-   Method and analysis

@@

There are several important concepts to consider when conducting longitudinal analyses in a developmental context. These include different ways of thinking about the developmental course, whether certain periods of development are relatively sensitive or insensitive to various types of insults or stressors, whether some time periods or situations inhibit the expression of individual differences due to extreme environmental pressures, and whether the same behavior manifested at different times represents the same or different phenomena.

@@

For example, collecting two or more within-person observations on the same construct at different times enables estimation of individual rates of change (slopes) where more observations allow for more precise estimates of individual slopes (random slopes), as well as characterization of non-linear development. Rate of change or other trajectory characteristics may be more informative about individuals than the simple snapshots of level differences that cross-sectional data are limited to informing about. Cross-sectional age-related differences across individuals are poor substitutes for longitudinal trajectory estimates, except under highly restrictive assumptions, e.g., parallel trajectories and lack of age, cohort and experience effects (@thompson2011). Appreciation of these and other issues can help to guide the analysis and interpretation of data and aid translation to clinical and public health applications.

@@

Adolescent development progresses normatively from less mature to more mature levels of functioning. However, unique epochs and experiences can alter the course of this idealized form of development. Consider research that shows cannabis use during adolescence is associated with later psychosis to a greater degree than cannabis use initiated later in development (@arseneault2002; @bechtold2016; @hasan2020; @semple2005). Similarly, rodent brains are especially sensitive to the neurotoxic effects of alcohol on brain structure and learning early in development, corresponding to early adolescence in humans (@spear2016; @crews2000; @ji2018). In another example, longitudinal data from the National Consortium on Alcohol and Neurodevelopment in Adolescence (NCANDA) show that binge drinking is associated more strongly with decrements in gray matter volume early in adolescence compared to later (@infante2022). These examples highlight the importance of considering the role of vulnerable periods – e.g., temporal windows of rapid brain development or remodeling during which the effects of environmental stimuli on the developing brain may be particularly pronounced– when trying to establish an accurate understanding of the association between exposures and outcomes.

@@

Whereas vulnerable periods heighten neurobiological susceptibility to environmental influences, at other times, environmental exposures will tend to suppress stability and disrupt the orderly stochastic process of normative development (e.g., @schulenberg2019). This situation reflects a developmental disturbance in that the normal course of development is “altered” for a time by some time-limited process. In such cases, we might find that prediction of behavior in the period of the disturbance is reduced and/or, similarly, the behavior exhibited during the disturbance might have less predictive power with respect to distal outcomes compared to the behavior exhibited before and following the disrupted period. That is, once the environmental pressures are removed (or the individual is removed from the environment), patterns of individual differences (and autoregressive effects) recover to levels similar to those prior to entering the environment.

@@

Normative development can also be upended by experiences (e.g., drug use) that, through various mechanisms, disrupt the normal flow of development wherein each stage establishes a platform for the next. For instance, substance use could lead to association with deviant peers, precluding opportunities for learning various adaptive skills and prosocial behaviors, in effect creating a “snare” that delays psychosocial development, such as maturing out of adolescent antisocial behavior (@moffitt2015). Relatedly, the consequences of these types of events can cascade (e.g., school dropout, involvement in the criminal justice system) so that the effects of the snare are amplified (e.g., @masten2005; @rogosch2010). Although conceptually distinct from vulnerable periods, both types of developmental considerations highlight the importance of viewing behavior in the context of development and attempting to determine how various developmental pathways unfold. Longitudinal data are crucial in this context to assess individual levels of development prior to and following onset of experiences or other environmental factors; e.g., the ABCD Study collected data starting at ages 9-10 and hence before the onset of substance use for the vast majority of participants.

### Mediational Processes.

Questions regarding the biological mechanisms whereby exposures impact outcomes can often be framed in terms of mediation analyses (@mackinnon2007; @vanderweele2016). Mediation analyses can be implemented using the causal steps approach (@baron1986) and structural equation models (SEM) (@preacher2011). More recently, mediation models have been adapted for longitudinal exposures, mediators, and/or outcomes (@bind2016; @vanderweele2017). All of these modeling approaches decompose the total effects of an exposure on an outcome into direct and indirect effects, where indirect effects of an exposure flow through its impact on a mediating process. @vanderweele2017 details conditions under which the direct and indirect causal effects can be in a longitudinal setting. An important example of mediational analyses in the ABCD Study is the impact of exposures on behavioral outcomes (e.g., neurocognition, mental health, substance use) via their impact on the brain, as quantified by imaging-derived phenotypes (IDPs). Methods for mediational analyses using multi-dimensional IDPs have been developed and applied to functional MRI data (@lindquist2012; @zhao2018).

@@

## Considerations and Challenges

The hallmark characteristic of longitudinal data analysis (LDA) is the administration of repeated measurements of the same constructs on assessment targets (e.g., individuals, families) across time. The primary rationale for collecting longitudinal data is to assess within-person change over time, allowing researchers to estimate individual developmental trajectories and the genetic and person-level factors that may impact these trajectories. Administering repeated measurements more frequently or over longer periods enables researchers to ask more nuanced questions and to make stronger inferences.

@@

Although the clear leap from cross-sectional to the realm of longitudinal data involves going from one assessment to two or more assessments, there are also notable distinctions in designs based on two-assessment points versus three or more measurement occasions. Just as cross-sectional data can be informative in some situations, two waves of data can be beneficial in contexts such as when an exposure is involved (e.g., pre/post tests), or if the central goal is prediction (e.g., trying to predict scores on Variable A at time T as a function of prior scores on Variable A and Variable B at time T-1). At the same time, analyses of data based on two assessments are inherently limited on multiple fronts. As @rogosa1982 noted over forty years ago, “Two waves of data are better than one, but maybe not much better” (p. 744).

These sentiments are reflected in more contemporary recommendations regarding best-practice guidelines for prospective data, which increasingly emphasize the benefits of additional measurement occasions for trajectory estimation, model identification and accurate parameter inferences. This is also consistent with recommendations that developmental studies include three or more assessment points, given it is impossible for data based on two-time points to determine the shape of development (given that linear change is the only estimable form for two assessment waves; (see @duncan2009). Research designs that include three (but preferably more) time points allow for non-linear trajectory estimation and increasingly nuanced analyses that more adequately tease apart sources of variation and covariation among the repeated assessments (@king2018)– a key aspect of developmental research.

To illustrate, developmental theories are useful for understanding patterns of within-individual change over time (discussed in further detail, below); however, two data points provide meager information on change at the person level. This point is further underscored in a recent review of statistical models commonly touted as distinguishing within-individual vs between-individual sources of variance in which the study authors concluded “… researchers are limited when attempting to differentiate these sources of variation in psychological phenomenon when using two waves of data” and perhaps more concerning, “…the models discussed here do not offer a feasible way to overcome these inherent limitations” (@littlefield2021). It is important to note, however, that despite the current focus on two-wave designs versus three or more assessment waves, garnering three assessment points is not a panacea for longitudinal modeling. Indeed, several contemporary longitudinal models designed to isolate within-individual variability (e.g., the Latent Curve Model with Structured Residuals [LCM: SR]; @curran2014) require at least four assessments to parameterize fully and, more generally, increasingly accurate and nuanced parameter estimates are obtained as more assessment occasions are used (@duncan2009).

@@

If one were to try to sum up what developmental trajectories in a living organism are exactly, one could plausibly argue they are the patterns of stability and change in its phenotypes as the organism traverses the life course. Symbolically, developmental trajectories can be expressed as fi(t), a possibly multivariate function of time t, specific to the ith individual and typically taking values in the real numbers for continuous phenotypes and the integers for discrete phenotypes. Ideally, t is a biologically meaningful temporal index (e.g., calendar age) as opposed to an exogenous progression of events (e.g., study visit number). Properties of interest might include rate of change over time, degree of smoothness (e.g., continuously differentiable), shape (e.g., polynomial or asymptotic behavior), how and how much f(t) differs across individuals, and what factors predict either within-individual variation (at different times) or between-individual variation (either overall or at specific times).

There are a few different ways to think about patterns of stability and change (see Figure 1). Consider measuring school disengagement at the start of middle school and the end of middle school. A common first step may be to compare sixth graders’ average disengagement values and eighth graders’ disengagement values. This comparison of the average scores for the same group of individuals at multiple time points is referred to as “mean-level”, as it provides information about change over time (or lack thereof) for an outcome of interest aggregated across members of a group. In contrast, “between-individual” stability could be assessed, e.g., by calculating the Spearman correlation between the values obtained at different time points (e.g., ‘disengagement in sixth grade’ with ’disengagement in eighth grade). This analysis focuses on the degree to which individuals retain their relative placement in a group across time. Consider someone who reported the lowest frequencies of disengagement in 6th grade and may report significantly higher disengagement over middle school (i.e., exhibit high levels of change), but report the lowest frequencies of disengagement in eighth grade. That is, the individual is manifesting rank-order stability, even in the context of high mean-level change.

Both types of stability and change are important. Mean-level change in certain traits might help to explain why, in general, populations of individuals tend to be particularly vulnerable to the effects of environmental factors in specific age ranges; rank-order stability might help to quantify the extent to which certain characteristics of the individual are more or less trait-like compared to others. For example, in some areas of development, considerable mean-level change occurs over time (e.g., changes in Big 5 personality traits (@bleidorn2022), but exhibit relatively high rank-order stability, at least over shorter measurement intervals (@bleidorn2022; @roberts2000; @roberts2006).

Despite the useful information afforded by examining mean-level and rank-order stability and change, these approaches are limited in that they provide little information about the overall patterns of within-individual change and, in turn, can result in fundamental misinterpretations about substantial or meaningful changes in an outcome of interest (@curran2011). For example, questions related to the impact of early-onset substance use on brain development focus on changes within a given individual (i.e., intraindividual differences). The ABCD Study will provide researchers with over ten time points for certain constructs (e.g., substance use) across a ten-year period, allowing for a detailed study of some within-person processes.

@@

### Use of appropriate longitudinal models
There is growing recognition that statistical models commonly applied to longitudinal data often fail to align with the developmental theory they are being used to assess (e.g., @curran2011; @hoffman2015; @littlefield2021). First, developmental studies typically involve the use of prospective data to inform theories that are concerned with clear within-person processes (e.g., how phenotypes change or remain stable within individuals over time, (e.g., @curran2011)). Despite this, methods generally unsuited for disaggregating between- and within-person effects (e.g., cross-lagged panel models [CLPM]) remain common within various extant literatures. Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time (see @littlefield2021; @orth2021). Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual’s mean (or trait) level across time (e.g., random-intercept cross-lagged panel model [RI-CLPM], @hamaker2015); latent curve models with structured residuals [LCM-SR], @curran2014). It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context (see @littlefield2021, for further discussion).

Second, many statistical models assume certain characteristics about the data to which they are being applied. Common assumptions of parametric statistical models (e.g., linear mixed-effects models) include normality and equality of variances. These assumptions should be carefully considered before finalizing analytical approaches, so that valid inferences can be made from the data, as violation of a model’s assumptions can substantively invalidate the interpretation of results. For example, longitudinal data can exhibit heterogeneous variability (i.e., the variance of the response changes over the duration of the study) that may need to be accounted for within a model. Another pertinent modeling assumption is whether trajectories are linear or non-linear. With two or three assessments per individual, usually only a linear model of within-person change is feasible.

As the study progresses and more time points are assessed, the potentially nonlinear aspects of trajectories can be assessed, for example using quadratic functions of time. Methods that make even fewer assumptions about trajectory shapes, such as nonparametric curve estimation at the mean (e.g., Generalized Additive Mixed Models [GAMMs]; (@wood2017)) and at the individual level (e.g., Functional Data Analysis [FDA]; @ramsay2002) may also become useful. Note, baseline age in the ABCD Study ranges over two full years; for some outcomes it may be feasible to include a possibly nonlinear effect of baseline age along with a linear effect of within-person change in age even with only two or three assessment times (@thompson2013).

### Continuous and Discrete Outcomes
Repeated assessments within the ABCD Study can be based on continuous or discrete measures. Examples of discrete measures include repeated assessments of binary variables (e.g., past 12-month alcohol use disorder status measured across ten years), ordinal variables (e.g., caregiver-reported items measuring emotional and behavioral concerns via the Child Behavior Checklist including the categories of “Not True”, “Somewhat True”, and “Very True”), and count variables (e.g., number of cigarettes smoked per day). In many ways, the distributional assumptions of indicators used in longitudinal designs mirror the decision points and considerations when delineating across different types of discrete outcome variables, a topic that spans entire textbooks (e.g., see @lenz2016). For example, the Mplus manual (@muthen2017) includes examples of a) censored and censored-inflated models, b) linear growth models for binary or ordinal variables, c) linear growth models for a count outcome assuming a Poisson model, and d) linear growth models for a count outcome assuming a zero-inflated Poisson model. Beyond these highlighted examples, other distributions (e.g., negative binomial) can be assumed for the indicators when modeling longitudinal data (@ren2022). These models account for issues that may occur when working with discrete outcomes, including overdispersion, i.e., when the variance is higher than would be expected based on a given parametric distribution (see @lenz2016). Given the sheer breadth of issues relevant to determining adequate models for discrete outcomes, it is not uncommon for texts on LDA to only cover models and approaches that assume continuous variables (e.g., @little2013). However, some textbooks on categorical data analysis provide more detailed coverage of the myriad issues and modeling choices to consider when working with discrete outcomes: @lenz2016, Chapter 11 for matched pair/two-assessment designs; Chapter 12 for marginal and transitional models for repeated designs, such as generalized estimating equations, and Chapter 13 for random effects models for discrete outcomes.

### Issues in attributing longitudinal change to development

Systematic changes over time in a variable of interest are not always attributable to development: various pitfalls with longitudinal data can complicate or even invalidate this conclusion. For example, if data missingness or participant dropout are related to the values of the outcome, changing sample composition as the study progresses can bias mean trajectory estimates (we describe this in more detail in Section 3.1.7 below). 
Another prerequisite for valid developmental interpretations of longitudinal data is to establish whether a construct is measured consistently over time, i.e., longitudinal measurement invariance (@liu2017; @van2015; @willoughby2012). Establishing longitudinal measurement invariance ensures that change over time for a given construct is attributable to individual development rather than merely a measurement artifact. For instance, one study using data from the ABCD Study (@brislin2023) found differential item functioning in two items from a brief delinquency measure, revealing significant bias in an arrest item across Black and White youth. More specifically, Black youth were more likely to report being arrested compared to White youth with similar levels of delinquency. Prevalence rates of delinquent behavior would have been severely biased if measurement invariance had not been tested. Alternatively, @vize2023 showed partially strong to strong evidence of longitudinal measurement invariance across broad externalizing dimensions in youth taking in the ABCD Study, suggesting that changes observed over time in these constructs were not due to systematic measurement error, but likely reflect true developmental change.
